On the number of tiles \touched{} by a\\ line segment on a rectangular grid

### 2021-08-21 paper.tex
Get rid of "\orient" command

\newcommand{\mr}{\mathrm}
\newcommand{\st}{\mid}%{\,|\,}
\newcommand{\cond}{\,|\,} % less spacing that \mid
\newcommand{\diff}{\,\mathrm d}
\let\Re\relax % remove the command...
\DeclareMathOperator{\Re}{Re} % ...to define it differently
\let\Im\relax
\DeclareMathOperator{\Im}{Im}
\DeclareMathOperator{\E}{E}
\newcommand{\funt}{\tau} % function that gives maximum number of touched tiles for real-valued length
\newcommand{\funl}{\lambda} % function that gives infimum real-valued length for a given number of touched tiles
\newcommand{\funti}{T} % function that gives maximum number of touched tiles for integer-valued length
\newcommand{\funli}{\Lambda} % function that gives minimum integer-valued length for a given number of touched tiles
\newcommand{\funta}{\varphi} % function that gives average number of touched tiles for real-valued length
\newcommand{\len}{\ell} % length variable, real-valued
\newcommand{\leni}{\ell} % length variable, integer-valued. Same notation, but I keep the different command just in case
\newcommand{\tiles}{t} % number of tiles (integer-valued)
\newcommand{\isolr}{i^+}
\newcommand{\jsolr}{j^+}
\newcommand{\isoli}{i^\ast}
\newcommand{\jsoli}{j^\ast}
\newcommand{\genvar}{r}
\newcommand{\orient}{\theta}
\newcommand{\mss}{M}
\newcommand{\touch}{visit}
\newcommand{\touches}{visits}
\newcommand{\touched}{visited}
\newcommand{\touching}{visiting}

@@

@@

Given $\len \in \mathbb R^+$, consider a segment of length $\len$ with random position and orientation. Specifically, the coordinates $x_1$, $y_1$ of the first endpoint are independent random variables uniformly distributed on $[0,a)$ and $[0,b)$ respectively. The orientation $\orient$ of the segment, defined as the angle with respect to the horizontal axis of the vector that goes from the first to the second endpoint, is uniformly distributed on $[0,2\pi)$. The variables $x_1$, $y_1$ and $\theta$ determine the coordinates $x_2$, $y_2$ of the second endpoint. Note that events of probability $0$, such as $x_1=a$ or the segment passing through a grid point, can be disregarded for computing the average number of tiles \touched{} by the segment.

@@

Given the coordinates $(x_1, y_1)$ of the first endpoint of the segment, with $0 \leq x_1 \leq a$, $0 \leq y_1 \leq b$, the second endpoint $(x_2, y_2)$ lies on a circle with radius $\len$ centered at $(x_1,y_1)$, as shown in Figure~\ref{fig: Pr_i}. The segment orientation is a random angle $\orient$ uniformly distributed on $[0,2\pi)$. It is clear from the figure that $i \geq n$ if and only if $x_2 \geq a(n-1)$ or $x_2 \leq -a(n-2)$; and these events are exclusive for $n \geq 2$. Thus
\begin{equation}
\label{eq: Pr i cond}
\Pr[i \geq n \cond x_1, y_1] = \Pr[x_2 \geq a(n-1) \cond x_1, y_1] + \Pr[x_2 \leq -a(n-2) \cond x_1, y_1].
\end{equation}
The two conditional probabilities on the right-hand side of \eqref{eq: Pr i cond} are different in general. However, averaging over $x_1, y_1$ gives, by symmetry, $\Pr[x_2 \geq a(n-1)] = \Pr[x_2 \leq -a(n-2)]$. In addition, the coordinate $y_1$ does not have any influence on these events, and therefore conditioning on $x_1, y_1$ is the same as conditioning on $x_1$. This implies that, for $n \geq 2$,
\begin{equation}
\label{eq: Pr i = 2 Pr}
\Pr[i \geq n] = 2 \Pr[x_2 \geq a(n-1)] = 2 \E[\Pr[x_2 \geq a(n-1) \cond x_1]].
\end{equation}

@@

Consider the event $x_2 \geq a(n-1)$ conditioned on $x_1$, with $n \geq 2$. There are three possibilities depending on $x_1$, $n$ and $\len$. If $a(n-1) < \len$, regardless of $x_1$ the length $\len$ is enough for $x_2$ to exceed $a(n-1)$ for some angles $\orient$. This is depicted in Figure~\ref{fig: Pr_i_full}, where the section of the arc with solid line represents those angles for which $x_2 > a(n-1)$. If $a(n-2) < \len \leq  a(n-1)$, the length will be enough provided that $x_1 \geq a(n-1)-\len$, which corresponds to the shaded region in Figure~\ref{fig: Pr_i_notfull}. Lastly, if $\len \leq  a(n-2)$ it is not possible for $x_2$ to exceed $a(n-1)$. Note that the coordinate $y_1$ is irrelevant to this.

>>>>

\newcommand{\mr}{\mathrm}
\newcommand{\st}{\mid}%{\,|\,}
\newcommand{\cond}{\,|\,} % less spacing that \mid
\newcommand{\diff}{\,\mathrm d}
\let\Re\relax % remove the command...
\DeclareMathOperator{\Re}{Re} % ...to define it differently
\let\Im\relax
\DeclareMathOperator{\Im}{Im}
\DeclareMathOperator{\E}{E}
\newcommand{\funt}{\tau} % function that gives maximum number of touched tiles for real-valued length
\newcommand{\funl}{\lambda} % function that gives infimum real-valued length for a given number of touched tiles
\newcommand{\funti}{T} % function that gives maximum number of touched tiles for integer-valued length
\newcommand{\funli}{\Lambda} % function that gives minimum integer-valued length for a given number of touched tiles
\newcommand{\funta}{\varphi} % function that gives average number of touched tiles for real-valued length
\newcommand{\len}{\ell} % length variable, real-valued
\newcommand{\leni}{\ell} % length variable, integer-valued. Same notation, but I keep the different command just in case
\newcommand{\tiles}{t} % number of tiles (integer-valued)
\newcommand{\isolr}{i^+}
\newcommand{\jsolr}{j^+}
\newcommand{\isoli}{i^\ast}
\newcommand{\jsoli}{j^\ast}
\newcommand{\genvar}{r}
\newcommand{\mss}{M}
\newcommand{\touch}{visit}
\newcommand{\touches}{visits}
\newcommand{\touched}{visited}
\newcommand{\touching}{visiting}

@@

From Theorem~\ref{theo: funt, sq, form} it is easily seen that odd values of $\funt(\len)$ correspond to $\isoli = \jsoli$, whereas even values are achieved with $\jsoli = \isoli+1$. Given $\tiles \geq 3$ with $t$ odd, $\funt(\len) = \tiles$ if and only
\begin{equation}
\label{eq: square, odd max tiles, lengths}
\frac \len a \in \left( \frac{\tiles-3}{\sqrt{2}}, \frac{ \sqrt{(\tiles^2-2)^2+1}}{\sqrt{2}} \right].
\end{equation}
For $\tiles \geq 3$, $t$ even, $\funt(\len) = \tiles$ if and only if
\begin{equation}
\label{eq: square, even max tiles, lengths}
\frac \len a \in \left(\frac{ \sqrt{(\tiles^2-3)^2+1}}{\sqrt{2}}, \frac{\tiles-2}{\sqrt{2}} \right].
\end{equation}

@@

Given $\len \in \mathbb R^+$, consider a segment of length $\len$ with random position and orientation. Specifically, the coordinates $x_1$, $y_1$ of the first endpoint are independent random variables uniformly distributed on $[0,a)$ and $[0,b)$ respectively. The orientation $\theta$ of the segment, defined as the angle with respect to the horizontal axis of the vector that goes from the first to the second endpoint, is uniformly distributed on $[0,2\pi)$. The variables $x_1$, $y_1$ and $\theta$ determine the coordinates $x_2$, $y_2$ of the second endpoint. Note that events of probability $0$, such as $x_1=a$ or the segment passing through a grid point, can be disregarded for computing the average number of tiles \touched{} by the segment.

@@

Given the coordinates $(x_1, y_1)$ of the first endpoint of the segment, with $0 \leq x_1 \leq a$, $0 \leq y_1 \leq b$, the second endpoint $(x_2, y_2)$ lies on a circle with radius $\len$ centered at $(x_1,y_1)$, as shown in Figure~\ref{fig: Pr_i}. The segment orientation is a random angle $\theta$ uniformly distributed on $[0,2\pi)$. It is clear from the figure that $i \geq n$ if and only if $x_2 \geq a(n-1)$ or $x_2 \leq -a(n-2)$; and these events are exclusive for $n \geq 2$. Thus
\begin{equation}
\label{eq: Pr i cond}
\Pr[i \geq n \cond x_1, y_1] = \Pr[x_2 \geq a(n-1) \cond x_1, y_1] + \Pr[x_2 \leq -a(n-2) \cond x_1, y_1].
\end{equation}
The two conditional probabilities on the right-hand side of \eqref{eq: Pr i cond} are different in general. However, averaging over $x_1, y_1$ gives, by symmetry, $\Pr[x_2 \geq a(n-1)] = \Pr[x_2 \leq -a(n-2)]$. In addition, the coordinate $y_1$ does not have any influence on these events, and therefore conditioning on $x_1, y_1$ is the same as conditioning on $x_1$. This implies that, for $n \geq 2$,
\begin{equation}
\label{eq: Pr i = 2 Pr}
\Pr[i \geq n] = 2 \Pr[x_2 \geq a(n-1)] = 2 \E[\Pr[x_2 \geq a(n-1) \cond x_1]].
\end{equation}

@@

Consider the event $x_2 \geq a(n-1)$ conditioned on $x_1$, with $n \geq 2$. There are three possibilities depending on $x_1$, $n$ and $\len$. If $a(n-1) < \len$, regardless of $x_1$ the length $\len$ is enough for $x_2$ to exceed $a(n-1)$ for some angles $\theta$. This is depicted in Figure~\ref{fig: Pr_i_full}, where the section of the arc with solid line represents those angles for which $x_2 > a(n-1)$. If $a(n-2) < \len \leq  a(n-1)$, the length will be enough provided that $x_1 \geq a(n-1)-\len$, which corresponds to the shaded region in Figure~\ref{fig: Pr_i_notfull}. Lastly, if $\len \leq  a(n-2)$ it is not possible for $x_2$ to exceed $a(n-1)$. Note that the coordinate $y_1$ is irrelevant to this.
